Gaussian Process regression for astronomical time-series

### 2022-09-15 src/tex/ms.tex
[showyourwork] overleaf sync

% Metadata Information
\jname{Xxxx. Xxx. Xxx. Xxx.}
\jvol{AA}
\jyear{2022}
\doi{10.1146/((please add article doi))}

@@

%Authors, affiliations address.
\author{Suzanne Aigrain,$^1$ and Daniel Foreman-Mackey$^2$
  \affil{$^1$Department of Physics, University of Oxford, Oxford, UK, OX1 3RH; email: suzanne.aigrain@physics.ox.ac.uk}
  \affil{$^2$Center for Computational Astrophysics, Flatiron Institute, New York, USA, NY 10010}}

%Abstract
\begin{abstract}
  Abstract text, approximately 150 words.
\end{abstract}

%Keywords, etc.
\begin{keywords}
  keywords, separated by comma, no full stop, lowercase
\end{keywords}
\maketitle

@@

Perhaps the earliest use of GPR in the refereed astronomical literature that will be familiar to a modern reader was published by \citet{prh92a} in the context of quasar variability, and for a long time this remained its main application domain in astronomy. GPR then gradually appeared in other areas, starting with photometric redshift estimation \citep{2006ApJ...647..102W}, then exoplanet transit observations \citep{2009ApJ...704...51C,2012MNRAS.419.2683G} and radial velocity planet searches \citep{2012MNRAS.419.3147A,2014MNRAS.443.2517H}. Nonetheless, GPR remained relatively niche and few astronomers had heard of it until a few years ago. To illustrate this point, we searched on the NASA Astrophysics Data System (ADS) for articles published in refereeed astronomy and astrophysics journals with the words ``Gaussian Process'' in full text of the article \autoref{fig:literature}.
After an increase in popularity throughout the 1990s, the use of GPs in astrophysics remained fairly constant around $\sim20$ publications per year until 2010.
Since 2010, the popularity of GPs has grown significantly, and in 2021, more than 500 refereed papers referencing GPs were published in the astrophysics literature.

@@

\begin{figure}[ht]
  \centering
  \script{literature.ipynb}
  \includegraphics[width=0.5\linewidth]{figures/literature.pdf}
  \caption{The number of referred publications in the astronomy and astrophysics literature that include the text ``Gaussian process'' as indexed by the NASA Astrophysics Data System (ADS). \suz{Add paper network?}}
  \label{fig:literature}
\end{figure}

@@

\subsubsection{Link to Gaussian Process Regression}\label{sec:gp-lsq-link}
% \dan{I think we should explicitly re-write this as $\mathcal{L}(\meanparams,\,\hyperparams)$ and indicate that $K$ and $m$ depend on these parameters. We should really highlight that this is a drop-in replacement for the likelihood defined in the previous section.} \suz{Is this ok?}
Let us now re-write the likelihood in matrix form:
\begin{equation}
\label{eq:gp_like}
  \mathcal{L}(\meanparams,\hyperparams) = 
  \frac{1}{\sqrt{2\,\pi\,|\mathbf{K}|}} \exp \left(-(\mathbf{y}-\mathbf{m})^{\mathrm{T}} \mathbf{K}^{-1}
  (\mathbf{y}-\mathbf{m}) \right),
\end{equation}
where, as before, the mean vector $\mathbf{m}$ has elements $m_i=m(t_i,\meanparams)$ and $\mathbf{K}$ is a purely diagonal ($N$, $N$) matrix with elements $K_{ij} = \delta_{ij} \sigma_i^2$ ($\delta_{ij}$ being the discrete Kronecker delta function). This is, of course, the \emph{covariance matrix} of the model. The notation $|\mathbf{A}|$ refers to the determinant of the matrix $\mathbf{A}$, and $\mathbf{A}^{-1}$ to its matrix inverse.

Now, instead of assuming that the covariance matrix takes this very specific form, let us allow a more flexible covariance model:
\begin{equation}
  K_{ij} = k(t_i,t_j,\hyperparams) + \delta_{ij} \sigma_i^2,
\end{equation}
where $k$ is a \emph{covariance function}, or \emph{kernel function}, controlled by parameters $\hyperparams$. The result is a GP, and its likelihood is still given by Equation~\ref{eq:gp_like}. Depending on the choice of kernel function and parameters, the covariance matrix can now have non-zero off-diagonal elements, allowing us to explicitly model correlated noise or stochastic signals in the data.
Note that the likelihood depends not only on the argument of the exponential, but also on the determinant of the covariance matrix, which therefore needs to be evaluated explicitly. This term acts as a built-in Occam's razor, automatically penalising more complex models.
%\dan{We should probably define the determinant notation here? I don't think we really clearly define our matrix notation anywhere else.}

@@

\subsubsection{The predictive equations}

@@

% \dan{More here.} \suz{There's examples of multi-variate GPs where time is only one of several inputs applied to modelling systematics in transit observations in \ref{sec:transit_fit} you can refer to here if you like. In particular I mention the ARD kernel there.}

\subsubsection{Physically motivated kernels}

Physical processes that are expected to be GPs appear frequently in astrophysics.
For example, a physical model that is specified by a parametric form for its ``power spectrum'' in the Fourier domain (think, the cosmic microwave background, or solar-like asteroseismic oscillations), could \emph{equivalently} be specified by its covariance function (or kernel function) in real space.
It is common practice in astrophysics to analyse datasets with these models by first Fourier transforming the data, and then fitting these transformed data using a physical model for the power spectrum.
There is a rich literature on this topic, discussing the motivation for this approach, and the rigorous propagation of measurement uncertainty for these datasets.
These analyses could also be performed using GPR without first transforming the data, but this has typically been avoided since na\"ive implementation of GPR would be significantly more computationally expensive, or even completely intractable.

There are, however, many cases where analysis in the time domain is preferred---at least in principle---over a Fourier analysis.
This is especially true when working with short, significantly unevenly sampled, or otherwise heterogeneous datasets, or when combined with non-stationary functions that cannot be compactly represented in the frequency domain. 
% \dan{More here or cut?} \suz{I think it's worth making the general point that in principle it's always better to work in the time domain when the sampling is irregular, rather than interpolate in order to compute a power spectrum? I'm also uncertain how well defined FTs of heteroskedastic datasets are?}

Another class of GPs that are commonly used for astrophysics are stochastic differential equations \citep[see][for a detailed discussion of these models]{Sarkka:2019}, although this terminology is not standard in the literature.
The most widely-used model in this class is the Ornstein--Uhlenbeck \citep{1930PhRv...36..823U}, or ``damped random walk'' process \citep[e.g.,][]{2009ApJ...698..895K, 2012A&A...546A..89B, 2020AJ....160..265H}.
These models---or the generalized ``continuous-time autoregressive moving average'' (CARMA) process \citep[e.g.,][]{2014ApJ...788...33K, celerite, 2022arXiv220108943Y}---are commonly used as a physically-motivated model for the stochastic time-variability of active galactic nuclei (AGN).
Although this fact is not always noted in the literature, these models are examples of GPs with specific kernel function choices.
Besides their physical interpretability, these models can also be evaluated with computationally efficient and scalable algorithms \citep[e.g.,][]{2014ApJ...788...33K, celerite}, which can be crucial for practical applications, as we discuss in \autoref{sec:fast}.
% \dan{Do we want to add some of the math here? It might be hard to make it useful without too many details.} \suz{If there's a didactic reference we can cite it might be better to do that. NB might want to cite the work of Caceres, Feigelson et al. using ARMA models for transit searches? \url{https://ui.adsabs.harvard.edu/abs/2019AJ....158...58C/abstract}}

@@

\dan{Say something here about assessing Gaussianity, and more generally model checking (rather than comparison; posterior predictive checks).}

@@

First, using the full dataset, we compute the Bayesian evidence integral for each of these model choices using a nested sampling algorithm implemented in the \textsf{jaxns} package \citep{jaxns}.
The Bayesian evidence integral is defined as:
\begin{equation}
  Z(H) \equiv p(\ydata\,|\,H) = \int p(\meanparams,\hyperparams\,|\,H)\,p(\ydata\,|\,\meanparams,\hyperparams,H)\,\mathrm{d}\meanparams\,\mathrm{d}\hyperparams
\end{equation}
where $H$ indicates the modeling choices, in this case the choice of kernel function.
For better or worse\footnote{A discussion of the relative merits of different model selection algorithms is beyond the scope of this review. \suz{Sure but can we link to a more in-depth discussion of the topic for interested readers?}}, the Bayesian evidence is frequently used in the astrophysics literature as an ingredient in model selection procedures.
Nested sampling is an algorithm for numerically estimating this integral, and we apply it to produce estimates of $Z$ for each choice of kernel $H$, and plot these results in the middle panel of \autoref{fig:assessment}.
In this figure, it is clear that the squared exponential and rational quadratic kernels are indistinguishable under this metric, while the Mat\'ern-3/2 kernel is somewhat disfavoured.

@@

@@

On the other hand, RV extraction pipelines routinely provide additional indicators measuring the characteristic width and asymmetry of the spectral lines, which are also affected by active regions, as well as spectral activity indicators which trace chromospheric emission in the cores of strong lines. \citet{2015MNRAS.452.2269R} developed a GP framework, schematically illustrated in \autoref{fig:RV_framework}, to model these activity indicators alongside the RVs, and to disentangle them from planetary signals. Each observed time-series is modelled as a linear combination of some latent variable $G(t)$, its time derivative $G'(t)$, white noise, and (in the case of the RVs only) one or more Keplerian signals. $G(t)$ does not have a direct physical interpretation; it loosely corresponds to $F^2$ in the $FF'$ framework of \citet{2012MNRAS.419.3147A}, so that $G'(t) \propto F(t) F'(t)$, but it is modelled as a (typically quasi-periodic) GP, which makes it possible to write down expressions for the covariance between any pair of observations from the any of the time-series included in the analysis. This results in a global covariance matrix of dimensions $(NM,NM)$, where $N$ is the number of observations and $M$ the number of activity indicators included in the analysis (including the RVs themselves). This framework has proved particularly successful for RV confirmation of transiting planets around young stars \citep{2019MNRAS.490..698B,2022MNRAS.512.3060Z}, enabling the detection of planetary signals almost 50 times smaller than the activity signals.  Open-source implementations of this framework have been published recently \citep{2022MNRAS.509..866B,2022A&A...659A.182D}, and extensions thereof are likely to play a significant role in the analysis of next regeneration RV surveys targeting Earth analogues, which need to achieve even better contrast between activity and planetary signals.

@@

\dan{Assigned: DFM}

@@

\dan{Assigned: DFM}

\suz{This section is assigned to DFM but as I started writing something about it for the introduction then decided to move it here I have left it in -- feel free to use this as a starting point or ignore it entirely!.}

The Active Galactic Nuclei (AGN) at the centre of many distant galaxies produce stochastic variability in their optical and radio emission, on timescales ranging from hours to years.

As discussed in Section~\ref{sec:intro}, this is --to the best of our knowledge-- the first area in which GPR makes an appearance in the refereed astronomical literature.

\suz{Refer to the first simulated example in \autoref{sec:quasar} here. Some of the text below may need to move to that Section.}

When the quasar is lensed by an intervening galaxy or galaxy cluster, multiple images can be formed, whose brightness can be monitored individually. Measurements of the time-delays between the resulting light curves can be used to constrain the Hubble constant, $H_0$ \citep{doi:10.1146/annurev.aa.30.090192.001523}. Early measurements of these time-delays in the decade following the discovery of the first multiply imaged quasar, the ``double quasar'' 0957$+$561, were hampered by the lack of a generative model for the light curves. Heuristic methods based on interpolation and cross-correlation were developed, but uncertainties were difficult to derive and results obtained with different instruments and different bandpasses weren't always consistent. \citet{prh92a} derived a simple $\chi^2$ metric under the assumption that the observed light curves are shifted, noisy versions of a single sample a Gaussian process with known covariance. Having proposed an empirical procedure to estimate the covariance function from the data, they then optimize the $\chi^2$ with respect to the time-delay between images, and apply standard methods used to derive an uncertainty. While their method relied on an adhoc method to estimate the covariance matrix, and did not propagate the corresponding errors into the final result, it did resolve the prior discrepancy between the estimates of the time-delay from optical and radio dataets \citet{prh92b}. Even then, the requirement to invert the covariance matrix was a barrier to the wider application of this method. This was addressed a few years later by \citet{pr95}, who introduced introduced a fast algorithm, to invert the covariance matrix for exponential kernels. Updated versions of this algorithm now form the basis for fast GP solvers discussed in Section~\ref{sec:fast}.

Quasar power spectrum estimation references: \citet{2014ApJ...788...33K, 2017MNRAS.470.3027K, 2019PASP..131f3001M, 2022ApJ...936..132Y}

\subsection{Compact objects, gravitational waves, and transients}
\suz{I thought this new heading would be a neat way to introduce applications to pulsars, non-quasar QPOs and GWs?}

% \subsubsection{Pulsars}

To the best of our knowledge, GPs were first applied to pulsar observations by \citet{2014PhRvD..90j4012V} in the context of pulsar timing studies aimed at detecting gravitational waves. They use a GP to model the noise on series of pulse timing measurements, which are used to search for spacetime distortions caused by the gravitational waves.

More recently, \citet{2016MNRAS.456.1374B,2019MNRAS.488.5702B,2022MNRAS.513.5861S} used GPs to model decades-long radio pulsar observations. In this case the GP is used to model the pulses themselves, and to track long-term changes in pulse periodicity and shape. This is done by evaluating (analytically) the first and second derivatives of the fitted GP and using those to find extrema and inflexion points of the pulse profile, along with the associated uncertainties.

% \subsubsection{Quasi-periodic oscillations}

Quasi-periodic oscillations: \suz{Came across this reference while looking for pulsar work: \citet{2022ApJ...936...17H}}

% \subsubsection{Gravitational waves}
Gravitational waves:
\begin{enumerate}
  \item In GW, they always work in the Fourier domain. Given infinite / periodic data, and stationary noise, this diagonalizes the covariance matrix \citep{2002PhRvD..65l2002A, 2017LRR....20....2R}. The periodicity is achieved by windowing the data then FFTing, but this breaks stationarity. This isn't a sensible assumption for the other examples we've discussed, and it gets tricky here for things like ringdown analysis \citep{2021arXiv210705609I}.
  \item Similarly they're always operating in the uniformly-sampled limit (Toeplitz matrices) \citep{2002PhRvD..65l2002A}.
  \item But the PSD of the noise is very complicated and well captured in the F domain. It would be non-trivial to implement it in the time domain and perhaps computationally intractable.
  \item Detailed discussion of approximating noise as diagonal in Fourier domain when all usual assumptions are not met \citep{UNSER1984231}.
\end{enumerate}

% \subsubsection{Transients}
Transients:
\begin{enumerate}
  \item Using GP interpolation: \citep{2016ApJS..225...31L, 2019AJ....158..257B, 2020ApJ...905...94V, 2021ApJS..255...24V}
\end{enumerate}

\dan{Point that this is the use of GPs for interpolation.}

>>>>

% Metadata Information
\jname{Annu. Rev. Astron. Astrophys.}
\jvol{AA}
\jyear{2022}
\doi{10.1146/TBD}

@@

%Authors, affiliations address.
\author{Suzanne Aigrain,$^1$ and Daniel Foreman-Mackey$^2$
  \affil{$^1$Department of Physics, University of Oxford, Oxford, UK, OX1 3RH; email: suzanne.aigrain@physics.ox.ac.uk}
  \affil{$^2$Center for Computational Astrophysics, Flatiron Institute, New York, USA, NY 10010; email: dforeman-mackey@flatironinstitute.org}}

%Abstract
\begin{abstract}
The last two decades have seen a major expansion in the availability, size, and precision of time-domain datasets in astronomy. Owing to their unique combination of flexibility, mathematical simplicity and comparative robustness, Gaussian Processes (GPs) have emerged recently as the solution of choice to model stochastic signals in such datasets. In this review we provide a brief introduction to the emergence of GPs in astronomy, present the underlying mathematical theory, and give practical advice considering the key modelling choices involved in GP regression. We then review applications of GPs to time-domain datasets in the astrophysical literature so far, from exoplanets to active galactic nuclei, showcasing the power and flexibility of the method. We provide worked examples using simulated data, with links to the source code, discuss the problem of computational cost and scalability, and give a snapshot of the current ecosystem of open source GP software packages.
Driven by further algorithmic and conceptual advances, we expect that GPs will continue to be an important tool for robust and interpretable time domain astronomy for many years to come.
\end{abstract}

%Keywords, etc.
\begin{keywords}
  Gaussian process regression, astronomy data analysis, time-series analysis, time domain astronomy, astrostatistics techniques, computational methods

\end{keywords}
\maketitle

@@

Perhaps the earliest use of GPR in the refereed astronomical literature that will be familiar to a modern reader was published by \citet{prh92a} in the context of quasar variability, and for a long time this remained its main application domain in astronomy. GPR then gradually appeared in other areas, starting with photometric redshift estimation \citep{2006ApJ...647..102W}, then exoplanet transit observations \citep{2009ApJ...704...51C,2012MNRAS.419.2683G} and radial velocity planet searches \citep{2012MNRAS.419.3147A,2014MNRAS.443.2517H}. Nonetheless, GPR remained relatively niche and few astronomers had heard of it until a few years ago. To illustrate this point, we searched on the NASA Astrophysics Data System (ADS) for articles published in refereeed astronomy and astrophysics journals with the words ``Gaussian Process'' in the full text of the article \autoref{fig:literature}.
After an increase in popularity throughout the 1990s, the use of GPs in astrophysics remained fairly constant around $\sim20$ publications per year until 2010.
Since 2010, the popularity of GPs has grown significantly, and in 2021, more than 500 refereed papers referencing GPs were published in the astrophysics literature. A snapshot of the areas where GPs are currently used in astronomy can be obtained readily from the keywords of those papers published in the last two years (using the NASA ADS analytics tools). This shows that approximately a third of these are cosmology papers, where GPs are primarily used as emulators; this falls outside the scope of this review since it doesn't involve time-series data. The second and largest grouping covers exoplanets and, by extension, stellar variability (since most exoplanet discoveries rely on indirect detection of the planetary signals in observations of their host stars). Finally, the remaining 30\% or so of the papers relate to other kinds of astrophysical variability including active galactic nuclei, compact objects, transients and gravitational waves.

@@

\begin{figure}[ht]
  \centering
  \script{literature.ipynb}
  \includegraphics[width=0.5\linewidth]{figures/literature.pdf}
  \caption{The number of referred publications in the astronomy and astrophysics literature that include the text ``Gaussian process'' as indexed by the NASA Astrophysics Data System (ADS). }
  \label{fig:literature}
\end{figure}

@@

\subsubsection{Link to Gaussian Process Regression}\label{sec:gp-lsq-link}
Let us now re-write the likelihood in matrix form:
\begin{equation}
\label{eq:gp_like}
  \mathcal{L}(\meanparams,\hyperparams) = 
  \frac{1}{\sqrt{2\,\pi\,|\mathbf{K}|}} \exp \left(-(\mathbf{y}-\mathbf{m})^{\mathrm{T}} \mathbf{K}^{-1}
  (\mathbf{y}-\mathbf{m}) \right),
\end{equation}
where, as before, the mean vector $\mathbf{m}$ has elements $m_i=m(t_i,\meanparams)$ and $\mathbf{K}$ is a purely diagonal ($N$, $N$) matrix with elements $K_{ij} = \delta_{ij} \sigma_i^2$ ($\delta_{ij}$ being the discrete Kronecker delta function). This is, of course, the \emph{covariance matrix} of the model. The notation $|\mathbf{A}|$ refers to the determinant of the matrix $\mathbf{A}$, and $\mathbf{A}^{-1}$ to its matrix inverse.

Now, instead of assuming that the covariance matrix takes this very specific form, let us allow a more flexible covariance model:
\begin{equation}
  K_{ij} = k(t_i,t_j,\hyperparams) + \delta_{ij} \sigma_i^2,
\end{equation}
where $k$ is a \emph{covariance function}, or \emph{kernel function}, controlled by parameters $\hyperparams$. The result is a GP, and its likelihood is still given by Equation~\ref{eq:gp_like}. Depending on the choice of kernel function and parameters, the covariance matrix can now have non-zero off-diagonal elements, allowing us to explicitly model correlated noise or stochastic signals in the data.
Note that the likelihood depends not only on the argument of the exponential, but also on the determinant of the covariance matrix, which therefore needs to be evaluated explicitly. This term acts as a built-in Occam's razor, automatically penalising more complex models.

@@

\subsubsection{The predictive equations}\label{sec:predeq}

@@

\subsubsection{Physically motivated kernels}
\label{sec:physical}
Physical processes that are expected to be GPs appear frequently in astrophysics.
For example, a physical model that is specified by a parametric form for its ``power spectrum'' in the Fourier domain (think, the cosmic microwave background, or solar-like asteroseismic oscillations), could \emph{equivalently} be specified by its covariance function (or kernel function) in real space.
It is common practice in astrophysics to analyse datasets with these models by first Fourier transforming the data, and then fitting these transformed data using a physical model for the power spectrum.
There is a rich literature on this topic, discussing the motivation for this approach, and the rigorous propagation of measurement uncertainty for these datasets.
These analyses could also be performed using GPR without first transforming the data, but this has typically been avoided since na\"ive implementation of GPR would be significantly more computationally expensive, or even completely intractable.

There are, however, many cases where analysis in the time domain is preferred---at least in principle---over a Fourier analysis.
This is especially true when working with short, significantly unevenly sampled, or otherwise heterogeneous datasets, or when combined with non-stationary functions that cannot be compactly represented in the frequency domain.

Another class of GPs that are commonly used for astrophysics are stochastic differential equations \citep[see][for a detailed discussion of these models]{Sarkka:2019}, although this terminology is not standard in the literature.
The most widely-used model in this class is the Ornstein--Uhlenbeck \citep{1930PhRv...36..823U}, or ``damped random walk'' process \citep[e.g.,][]{2009ApJ...698..895K, 2012A&A...546A..89B, 2020AJ....160..265H}.
These models---or the generalized ``continuous-time autoregressive moving average'' (CARMA) process \citep[e.g.,][]{2014ApJ...788...33K, celerite, 2022arXiv220108943Y}---are commonly used as a physically-motivated model for the stochastic time-variability of active galactic nuclei (AGN).
Although this fact is not always noted in the literature, these models are examples of GPs with specific kernel function choices.
Besides their physical interpretability, these models can also be evaluated with computationally efficient and scalable algorithms \citep[e.g.,][]{2014ApJ...788...33K, celerite}, which can be crucial for practical applications, as we discuss in \autoref{sec:fast}.

@@

@@

First, using the full dataset, we compute the Bayesian evidence integral for each of these model choices using a nested sampling algorithm implemented in the \textsf{jaxns} package \citep{jaxns}.
The Bayesian evidence integral is defined as:
\begin{equation}
  Z(H) \equiv p(\ydata\,|\,H) = \int p(\meanparams,\hyperparams\,|\,H)\,p(\ydata\,|\,\meanparams,\hyperparams,H)\,\mathrm{d}\meanparams\,\mathrm{d}\hyperparams
\end{equation}
where $H$ indicates the modeling choices, in this case the choice of kernel function.
The Bayesian evidence is frequently used in the astrophysics literature as an ingredient in model selection procedures.
Nested sampling is an algorithm for numerically estimating this integral, and we apply it to produce estimates of $Z$ for each choice of kernel $H$, and plot these results in the middle panel of \autoref{fig:assessment}.
In this figure, it is clear that the squared exponential and rational quadratic kernels are indistinguishable under this metric, while the Mat\'ern-3/2 kernel is somewhat disfavoured.

@@

Besides the model comparison tests discussed so far, it can also be useful to assess the quality of fit in a more absolute sense.
Like with the other assessment tools, there isn't anything particularly unique about GP models when it comes to model checking, although it is generally non-trivial to quantify goodness of fit metrics in all but the simplest cases \citep[e.g.,][]{gelman1995bayesian}.
Our preferred approach to model checking is to use graphical posterior predictive checks \citep[see][for example]{bayes-workflow}.
The idea behind posterior predictive checks is that the observed data $\ydata$ should not be an extreme outlier with respect to synthetic data $\ydata_\star$ sampled from the posterior:
\begin{equation}
  \ydata_\star \sim p(\ydata_\star\,|\,\ydata) = \int p(\ydata_\star\,|\,\meanparams,\hyperparams)\,p(\meanparams,\hyperparams\,|\,\ydata)\,\mathrm{d}\meanparams\,\mathrm{d}\hyperparams \quad.
\end{equation}
Operationally, a single sample of $\ydata_\star$ is generated by drawing a sample of $\meanparams$ and $\hyperparams$ from the posterior, and then drawing $\ydata_\star$ from the sampling distribution $p(\ydata_\star\,|\,\meanparams,\hyperparams)$ conditioned on the sampled values of $\meanparams$ and $\hyperparams$.
Datasets in astrophysics are typically sufficiently high dimensional that it is not feasible to visualize the full distribution of $\ydata_\star$, and we instead plot scalar statistics of the data, such as the mean, variance, or skew.
If the model is appropriate for the problem at hand, the same statistic computed on the observed data should be consistent with the posterior predictive distribution.

@@

On the other hand, RV extraction pipelines routinely provide additional indicators measuring the characteristic width and asymmetry of the spectral lines, which are also affected by active regions, as well as spectral activity indicators which trace chromospheric emission in the cores of strong lines. \citet{2015MNRAS.452.2269R} developed a GP framework, schematically illustrated in \autoref{fig:RV_framework}, to model these activity indicators alongside the RVs, and to disentangle them from planetary signals. Each observed time-series is modelled as a linear combination of some latent variable $G(t)$, its time derivative $G'(t)$, white noise, and (in the case of the RVs only) one or more Keplerian signals. $G(t)$ does not have a direct physical interpretation; it loosely corresponds to $F^2$ in the $FF'$ framework of \citet{2012MNRAS.419.3147A}, so that $G'(t) \propto F(t) F'(t)$, but it is modelled as a (typically quasi-periodic) GP, which makes it possible to write down expressions for the covariance between any pair of observations from the any of the time-series included in the analysis. This results in a global covariance matrix of dimensions $(NM,NM)$, where $N$ is the number of observations and $M$ the number of activity indicators included in the analysis (including the RVs themselves). This framework has proved particularly successful for RV confirmation of transiting planets around young stars \citep{2019MNRAS.490..698B,2022MNRAS.512.3060Z}, enabling the detection of planetary signals almost 50 times smaller than the activity signals.  Open-source implementations of this framework have been published recently \citep{2022MNRAS.509..866B,2022A&A...659A.182D}, and extensions thereof are likely to play a significant role in the analysis of next regeneration RV surveys targeting Earth analogues, which need to achieve even better contrast between activity and planetary signals. Very recently,  \citet{2022arXiv220506627C} proposed a GP regression network model for stellar activity signals in RV, showing that it can be used to model RVs of the Sun-as-a-star simultaneously with other activity indicators.

@@

As discussed in Section~\ref{sec:physical}, asteroseismology is an area where physically motivated GP models arise quite naturally. This has motivated a studies where GPs have been used to model stellar oscillations in the time-domain \citep{2017AJ....154..254G, 2018ApJ...865L..20F} but these remain very much an exception to the much more widespread practice of transforming the data into the Fourier domain, and identifying and modelling oscillation modes therein. Using a GP in this context makes particular sense when the data are irregularly sampled or in homogeneous, or when the GP is combined with a mean function that is much easier to specify in the time-domain (for example a planetary transit). For example, the \textit{PLATO} mission \citep{2014ExA....38..249R} aims to characterise all the host stars of the transiting planets it will discover through asteroseismology. Current plans for the \textit{PLATO} pipeline involve either subtracting the best-fit transit model from the light curve, or cutting and interpolating over the transits, before performing the seismic analysis. However, a GP model would  enable a simultaneous analysis, which should in principle be more robust. Conversely, \citet{2020A&A...634A..75B} demonstrated the importance of 
explicitly accounting for short-term stellar variability signals, including both granulation and oscillations, to obtain robust and accurate values of the transiting planets' parameters; and compared a range of GP models with different kernels for this purpose.

@@

The Active Galactic Nuclei (AGN) at the centre of many distant galaxies produce stochastic variability in their optical and radio emission, on timescales ranging from hours to years.
GPs can be useful descriptive models for these processes, and research in this domain has driven the development of new methods for GPR with large datasets.
In particular, as discussed in \autoref{sec:intro}, a GP was used by \citet{prh92a} to model the underlying variability of gravitationally lensed quasar 0957+561 to measure its time delay, in one of the first explicit uses of GPR in the refereed astronomical literature.

When the quasar is lensed by an intervening galaxy or galaxy cluster, multiple images can be formed, whose brightness can be monitored individually.
Measurements of the time-delays between the resulting light curves can be used to constrain the Hubble constant, $H_0$ \citep{doi:10.1146/annurev.aa.30.090192.001523}.
Early measurements of these time-delays in the decade following the discovery of the first multiply imaged quasar, the ``double quasar'' 0957$+$561, were hampered by the lack of a generative model for the light curves.
Heuristic methods based on interpolation and cross-correlation were developed, but uncertainties were difficult to derive and results obtained with different instruments and different bandpasses weren't always consistent.
\citet{prh92a} derived a simple $\chi^2$ metric under the assumption that the observed light curves are shifted, noisy versions of a single sample a Gaussian process with known covariance.
Having proposed an empirical procedure to estimate the covariance function from the data, they then optimized the $\chi^2$ with respect to the time-delay between images, and applied standard methods to derive an uncertainty.
While their method relied on an adhoc method to estimate the covariance matrix, and did not propagate the corresponding errors into the final result, it did resolve the prior discrepancy between the estimates of the time-delay from optical and radio datasets \citep{prh92b}.
Even then, the requirement to invert the covariance matrix was a barrier to the wider application of this method.
This was addressed a few years later by \citet{pr95}, who introduced a fast algorithm, to invert the covariance matrix for exponential kernels.
Updated versions of this algorithm now form the basis for fast GP solvers discussed in \autoref{sec:fast}.

GPR is also used to study the population-level properties of AGN samples, by characterizing and interpreting the statistics of the light curve variability for large samples.
The idea is that the properties of the stochastic variations of an AGN light curve are driven by an accretion disk near the central supermassive black hole, and that this variability provides an observational probe of the small scale astrophysical processes.
The standard approach is to fit a GP model to the light curve---typically using a ``damped random walk'' kernel, or the more general ``continuous-time autoregressive moving average'' (CARMA) kernel \citep[e.g.,][]{2014ApJ...788...33K}, since these permit efficient computations even with large datasets---and then to use the inferred hyper-parameter values to constrain the variability amplitudes and timescales of the system \citep[e.g.,][]{2014ApJ...788...33K, 2017MNRAS.470.3027K, 2019PASP..131f3001M, 2022MNRAS.514..164S}.
These amplitudes and timescales have been shown to empirically correlate with the fundamental properties of the AGN system, such as the mass of the central black hole, Eddington ratio, and bolometric luminosity \citep{2022ApJ...936..132Y}.

Driven by the success of this approach, and the scale of upcoming survey datasets from, for example, the Vera C. Rubin Observatory and the LSST survey, there has been an concerted effort within this field to develop GP methods that can scale to large datasets.
This began, perhaps, with \citet{pr95}, and the development of a fast GP solver for exponential kernels.
While widely used, this kernel is very restrictive, and more flexible models are required to capture the full range of variability observed in AGN light curves.
To this end, \citep{2014ApJ...788...33K} developed a scalable GP solver for the CARMA kernel, which uses a Kalman filtering approach to achieve linear scaling of the computational cost with the number of data points, which we discuss further in \autoref{sec:fast}.

\subsection{Compact objects, gravitational waves, and transients}

GPs have also been used for the analysis of a broad range of other astrophysical objects, including pulsars, gravitational wave sources, and other transients.
In this section, we highlight some of these remaining applications, in somewhat less detail than the other applications since we have less familiarity with these topics.
In many of these domains, it is standard practice to work in the Fourier domain by computing the fast Fourier transform (FFT) of the data, and then assume that the noise process is uncorrelated between frequency bins \citep[e.g.,][]{2017LRR....20....2R}.
In other words, the data covariance matrix in the Fourier domain is assumed to be diagonal, and this significantly reduces the computational cost of likelihood evaluations compared to a GP model with a dense covariance matrix.
This is a reasonable assumption for datasets with long observational baselines relative to the signals of interest, and stationary noise processes \citep[e.g.,][]{UNSER1984231}.
While these requirements are not met by many of the analyses that we have discussed, large datasets from pulsar timing surveys and gravitational wave interferometers do often satisfy these constraints, or can be transformed in such a way that they do.

\subsubsection{Pulsar timing} GPs are widely used in the analysis of pulsar observations, as a model of timing residuals, pulse shape variations, and the stochastic signals of interest.
In early work, \citet{2014PhRvD..90j4012V} describe a comprehensive set of GP models, and computational techniques for their implementation, for pulsar timing studies aimed at detecting gravitational waves.
These models include the effects of timing noise, dispersion-measure variations, and the stochastic gravitational wave background.
Such GP models are now routinely used for pulsar timing studies \citep[e.g.][]{2019MNRAS.489.3810P, 2020MNRAS.498.6044C, 2022MNRAS.510.4873A}.

GPs have also been used to model pulse shape variations in decades-long pulsar observations \citep{2016MNRAS.456.1374B,2019MNRAS.488.5702B,2022MNRAS.513.5861S}.
In this case, the GP is used to model the pulses themselves, and to track long-term changes in pulse periodicity and shape.
This is done by evaluating (analytically) the first and second derivatives of the fitted GP and using those to find extrema and inflection points of the pulse profile, along with the associated uncertainties.
Finally, \citet{2021MNRAS.508.4249P} used GPs to characterise the off-pulse noise background in observations of pulsars with large duty cycles and no obvious flat baseline.

\subsubsection{Gravitational waves} Within the LIGO--Virgo project, GP models are used to model the detector noise \citep[e.g.][]{2020CQGra..37e5002A}, and waveform modeling uncertainties \citep{2016PhRvD..93f4001M}.
For these datasets, the power spectrum has been empirically calibrated, and it includes non-trivial structure at all frequencies.
As discussed above, under some generally reasonable conditions, this noise process diagonalizes in the Fourier domain, and this empirically calibrated model can be used to compute the likelihood of a signal model also represented in the Fourier domain.
This type of model becomes significantly more complicated for ring-down analyses where edge effects invalidate the assumption that the data are periodic \citep{2021arXiv210705609I}.

\subsubsection{Transient classification} In the context of transient light curve classification, GPs have been widely used as a tool for interpolating sparsely-sampled and noisy photometric time series onto uniformly sampled grids that are more amenable to standard machine learning classification methods \citep[e.g.][]{2016ApJS..225...31L, 2019AJ....158..257B, 2020ApJ...905...94V}.
These pipelines differ in how they treat the covariances between photometric bands, with most learning a correlation length in effective wavelength.
Recently, \citet{2021ApJS..255...24V} proposed a wavelength covariance matrix that uses a distance metric defined by the throughput overlap between each pair of filters.
So far in this domain, GPs have primarily been used in a pre-processing step to produce data correctly structured for standard machine learning methods, but recent developments that integrate GPs with deep learning frameworks could permit these GPs to be embedded within the larger classification pipeline.

\subsubsection{Quasi-periodic oscillations}

One very active area of research where GPs have not been widely applied to date, is the characterisation of Quasi-Periodic Oscillations (QPOs). QPOs are seen in astrophysical transients right across the electromagnetic spectrum, including Gamma-ray bursts, X-ray binaries, flares in Sun-like stars and in magnetised pulsars, and fast radio bursts. Most QPOs are believed to arise from the interaction between one or more compact objects and an accretion disk, but a detailed understanding of the origin of most classes of QPOs remains elusive. Their phenomenology is typically studied in the time domain, by fitting the power spectrum of individual, continguous blocks of observations with empirical models, most frequently a sum of Lorentzian functions \citep[see][and references therein]{ingram_motta_QPOs}. Very recently, two studies \citep{Yang_2021,2022ApJ...936...17H} used GPs to model QPOs directly in the time-domain, with encouraging results both for the detection of QPOs in the presence of correlated noise and the characterisation of their properties such as frequency.
